Kaggle Titanic
--------------

Remove all variables, set working directory and seed and load required libraries.

```{r}
rm(list = ls())
set.seed(1234)
setwd('/home/greg/Dropbox/titanic')
library(tree)
library(ggplot2)
library(plyr)
```

Load training and test data.

```{r}
train <- read.csv('train.csv', stringsAsFactors = FALSE)
test <- read.csv('test.csv', stringsAsFactors = FALSE)
```

Notice that Survived is valid for all training data.

```{r}
sum(!train$Survived %in% c(0, 1))
```

## Cleaning

Combine training and test data together in preparation for cleaning and lowercase column names.

```{r}
full <- join(train, test, type = 'full')
colnames(full) <- tolower(colnames(full))
```

### Name

Extract title from names. Notice no missing titles.

```{r}
full$title <- NA
full[grepl(' Mr\\.', full$name), 'title'] <- 'mr'
full[grepl(' Mrs\\.', full$name), 'title'] <- 'mrs'
full[grepl(' Rev\\.', full$name), 'title'] <- 'rev'
full[grepl(' Dr\\.', full$name), 'title'] <- 'dr'
full[grepl(' Master\\.', full$name), 'title'] <- 'master'
full[grepl(' Miss\\.', full$name), 'title'] <- 'miss'
full[grepl(' Don\\.', full$name), 'title'] <- 'don'
full[grepl(' Dona\\.', full$name), 'title'] <- 'dona'
full[grepl(' Mme\\.', full$name), 'title'] <- 'mme'
full[grepl(' Ms\\.', full$name), 'title'] <- 'ms'
full[grepl(' Major\\.', full$name), 'title'] <- 'major'
full[grepl(' Lady\\.', full$name), 'title'] <- 'lady'
full[grepl(' Sir\\.', full$name), 'title'] <- 'sir'
full[grepl(' Mlle\\.', full$name), 'title'] <- 'mlle'
full[grepl(' Col\\.', full$name), 'title'] <- 'col'
full[grepl(' Capt\\.', full$name), 'title'] <- 'capt'
full[grepl(' Countess\\.', full$name), 'title'] <- 'countess'
full[grepl(' Jonkheer\\.', full$name), 'title'] <- 'jonkheer'
sum(is.na(full$title))

train <- subset(full, !is.na(survived))
ggplot(train, aes(title, survived)) + stat_summary(fun.y = mean, geom = 'bar')
ggplot(train, aes(title, survived)) + stat_summary(fun.y = sum, geom = 'bar')
```

Extract last name. Notice no missing last names.

```{r}
full$last <- sapply(strsplit(full$name, ','), function(x) x[1])
sum(is.na(full$last) || full$last == '')
```

Extract last.any.lived.

```{r}
full$last.any.lived <- sapply(1 : nrow(full), function(i) {
  survivors <- subset(
    full, last == full[i, 'last'] & 
    name != full[i, 'name'] & 
    !is.na(full[i, 'survived'])
  )
  ifelse (nrow(survivors) == 0, 0.5, 1.0 * (sum(survivors$survived) > 0))
})
```

### Age

Round age.

```{r}
full$age <- round(full$age)
```

`r sum(is.na(full$age))` missing ages.

```{r}
sum(is.na(full$age))
```

### Sex

No missing sex.

```{r}
sum(is.na(full$sex) || full$sex == ' ')
```

Factorize sex.

```{r}
full$sex <- factor(toupper(substring(full$sex, 1, 1)))
```

Explore sex survival.

```{r}
train <- subset(full, !is.na(survived))
ggplot(train, aes(sex, survived)) + stat_summary(fun.y = mean, geom = 'bar')
```

### Cabin

Explore cabins.

```{r}
full[order(full$cabin), c('last', 'survived', 'cabin', 'fare')]
```

Factorize cabin. Use pclass for missing cabins.

```{r}
full$cabin[full$cabin == ''] <- full$pclass[full$cabin == '']
full$cabin <- factor(sapply(full$cabin, function(cabin) substring(cabin, 1, 1)))
```

### Fare

One missing fare.

```{r}
sum(is.na(full$fare))
```

Round fare. 

```{r}
full$fare <- round(full$fare)
```

Explore fares.

```{r}
full[order(full$fare), 'fare']
```

```{r}
qplot(fare, data = full)
```


Notice some high fares. Even in third class. We need to determine fare per person.

```{r}
ggplot(full, aes(fare)) + geom_histogram(binwidth = 20) + facet_wrap(~ pclass)
```

### Ticket

Explore tickets.

```{r}
full[order(full$ticket), 'ticket']
```

Factorize ticket.

```{r}
full$ticket <- factor(sapply(full$ticket, function(ticket) {
  if (grepl('paris', ticket, ignore.case = TRUE)) {
    return('paris')
  } else if (substring(ticket, 1, 2) == 'PC') {
    return('pc')
  } else if (substring(ticket, 1, 2) == 'PP') {
    return('pp')
  } else if (substring(ticket, 1, 4) == 'P/PP') {
    return('ppp')
  } else if (substring(ticket, 1, 5) == 'S.O.C') {
    return('soc')
  } else if (substring(ticket, 1, 4) == 'SO/C') {
    return('soc')
  } else if (substring(ticket, 1, 4) == 'STON') {
    return('ston')
  } else if (substring(ticket, 1, 2) == 'XX') {
    return('xx')
  } else {
    return('xx')
  }
}))
```

### Embarked
`r sum(full$embarked == '')` missing embarked.

```{r}
sum(full$embarked == '')
ggplot(full, aes(embarked)) + geom_histogram()
full[full$embarked == '', 'embarked'] <- 'S'
ggplot(full, aes(embarked)) + geom_histogram()

train <- subset(full, !is.na(survived))
ggplot(train, aes(embarked, survived)) + stat_summary(fun.y = mean, geom = 'bar')
last_plot() + facet_wrap(~ sex)
```

Factorize embarked.

```{r}
full$embarked <- factor(full$embarked)
```

### Class

Explore how survival is related to passenger class.

First class passengers are more likely to live. Third class passengers are more likely to die.

```{r}
train <- subset(full, !is.na(survived))
ggplot(train, aes(pclass, survived)) + stat_summary(fun.y = mean, geom = 'bar') + facet_wrap(embarked ~ sex)
```


## Random Forest

```{r}
library(randomForest)

model.create <- function(data) {
  randomForest(survived ~ sex + pclass + cabin + embarked, data = data)
}

model.predict <- function(model, data) {predict(model, data)}
```

## SVM

```{r}
library(kernlab)

model.create <- function(data) {
  ksvm(survived ~ sex + pclass + cabin + embarked, data = data)
}

model.predict <- function(model, data) {predict(model, data, type = 'response')}
```

## Neural Net

```{r}
library(nnet)

model.create <- function(data) {
  nnet(survived ~ sex + pclass, data = data, size = 1, linout = FALSE, maxit = 10000)
}

model.predict <- function(model, data) {predict(model, data, type = 'raw')}
```

## Decision Tree

```{r}
model.create <- function(data) {
  tree(survived ~ sex + pclass + title + embarked, data = data)
}

model.predict <- function(model, data) {predict(model, data)}
```

## Cross Validate

```{r}
train <- subset(full, !is.na(survived))
row.names(train) <- NULL
n <- nrow(train)
best.score <- 0
best.model <- NULL

results <- data.frame(scores = replicate(30, {
  trial <- sample(1 : n, n * 4 / 5)
  model <- model.create(train[trial, ])
  cv <- train[-trial, ]
  score <- mean(cv$survived == round(model.predict(model, cv)))
  if (score > best.score) {
    best.score <<- score
    best.model <<- model
  }
  score
}))  

ggplot(results, aes(scores)) + geom_histogram(binwidth = .01)
```

## Submit

```{r}
test <- subset(full, is.na(survived))
row.names(test) <- NULL
answer <- data.frame(PassengerId = test$passengerid, Survived = round(predict(best.model, test)))
write.csv(answer, 'submit-002.csv', quote = FALSE, row.names = FALSE)
sum(read.csv('submit-001.csv') != answer)
```
